A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2025-05-01 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: TOC skeleton

@@

The insight is that probability distributions for a quantity are not defined by statistical mixtures of perfect preparations (i.e. convex combinations of extreme points), but by assigning a probability over each experimentally verifiable (i.e. open) interval of the possible values of said quantity (its spectrum). The two coincide in classical discrete spaces, which are the least physically interesting case as they do not even allow classical particle mechanics. Equivalently, \textbf{mixtures are about preparations, not measurements, and probability distributions are about measurements, not preparations}. In quantum mechanics, an equal mixture of spin up and spin left, for example, does not correspond to a 50\% probability of measuring spin up: the mixing coefficients and the outcome probability are different.

@@

We have clarified that mixing coefficients are not the same as probabilities of outcomes. Unlike most of the literature on probability, we will concentrate on mixtures and, particularly, on the following question: given an ensemble $\ens \in \Ens$ and a (Borel) set of ensembles $A \subseteq \Ens$, what fraction of $\ens$ can be constructed by a mixture of $A$? How much of $\ens$ can be explained as coming from preparations corresponding to $A$? For example, let $\Ens$ be the space of all probability distributions for a six-sided die. Let $\ens_{123456}$ be the uniform distribution over all outcomes. Let $A_1 = \{\ens_1\}$ where $\ens_1$ represents outcome one with 100\% probability. Since we can write $\ens_{123456} = \frac{1}{6} \ens_{1} + \frac{5}{6} \ens_{23456}$, where $\ens_{23456}$ represents a uniform distributions over the five outcomes, $1/6$ of $\ens$ can be constructed from $A$, but no more. Similarly, if $A_{12} = \{\ens_{1},\ens_{2}\}$, we can write $\ens = \frac{1}{3} \left(\frac{1}{2} \ens_1 + \frac{1}{2} \ens_2 \right)  + \frac{2}{3} \ens_{3456}$, so $1/3$ of $\ens$ can be constructed from $A$, but no more. Note that, given the uniform distribution, $1/6$ is exactly the probably for the event $A_1$ and $1/3$ the probability for event $A_{12}$. This gives the basic insight for our definition.

@@

One can then that the fraction capacity $\frcap_{\ens} : \Sigma_{\Ens} \to [0,1]$ is a set function that satisfies the following:
\begin{enumerate}
	\item non-negative and unit bounded - $0 \leq \frcap_{\ens}(A) \leq 1$
	\item monotone - $A \subseteq B \implies \frcap_{\ens}(A) \leq \frcap_{\ens}(B)$
	\item sub-additive - $\frcap_{\ens}(A \cup B) \leq \frcap_{\ens}(A) + \frcap_{\ens}(B)$
	\item continuous from below and above - $\frcap_{\ens}(\lim\limits_{i \to \infty} A_i) = \lim\limits_{i \to \infty} \frcap_{\ens}(A_i)$ for any increasing or decreasing sequence $\{A_i\}$.
\end{enumerate}

@@

>>>>

\section{TOC}

Overall goal: to create a generalization of probability theory and measure theory for general physical systems.

General question: how do we represent physical states? Different types of representations: talk about examples, ....

-- General problem

For this work: assume we have a locally convex topological vector space V. E is a bounded convex subset of V. Equipped with an entropy. Topology generated by entropic balls?

Separateness and orthogonality

Statistical quantities

Show that they reduce to standard things in classical mechanics and quantum mechanics.

-- Connection to Choquet theory

Choquet theory - in the case that the closure of the space is compact -> representation through measures. May not be unique

Choquet simplex. Uniqueness of decomposition and separateness/orthogonality.

-- Connection to signed probability

TODO: literature search on use of signed measure in QM. ( e.g. \href{https://arxiv.org/pdf/2302.00118}{this paper})

Characterize signed measures as an affine representation of the ensemble space (i.e. bounded convex subset of a locally convex topological vector space).

Show that the probability is signed if and only if multiple decomposition, non-additive, no Choquet simplex

Signed probability is unique, but does not characterize the convex set (ensemble space). Note: the other measures are defined on the extreme points. Knowing the extreme points means you know the convex space (i.e. take the hull). The non-additive measure does not characterize the extreme points, so it doesn't really characterize the space.

-- Our non-additive measure approach

State capacity

Fraction capacity. Show uniqueness. Also show uniqueness on extreme points.

Fraction capacity as supremum of all possible measures

Additivity of fraction capacity (and state capacity).

Single decomposition and classical spaces and measurement contexts

Show that they reduce to standard things in classical mechanics and quantum mechanics.

-- Connection to fuzzy logic?

Literature search.

Possible connection to
* Choquet integral
* Sugeno integral

? Space of fuzzy measure like we have a space of additive measures?

@@

The key insight is that \textbf{mixing coefficients and probabilities are different in general} and they coincides only when a particular ensemble can be understood as a statistical mixtures of perfect preparations of a given quantity (i.e. of orthogonal ensembles). This always happens in the classical case and over measurement contexts in the quantum case, but is not the general case. For example, an equal mixture of spin up and spin left, for example, does not correspond to a 50\% probability of measuring spin up. The key to develop a general theory of probability, then, is to characterize the mixing coefficients first and understand when they recover probabilities of outcomes.

@@

Given that mixing coefficients have a more general applicability than probability of outcomes, let us concentrate on the following problem: given an ensemble $\ens \in \Ens$ and a (Borel) set of ensembles $A \subseteq \Ens$, what fraction of $\ens$ can be constructed by a mixture of $A$? How much of $\ens$ can be explained as coming from preparations corresponding to $A$? For example, let $\Ens$ be the space of all probability distributions for a six-sided die. Let $\ens_{123456}$ be the uniform distribution over all outcomes. Let $A_1 = \{\ens_1\}$ where $\ens_1$ represents outcome one with 100\% probability. Since we can write $\ens_{123456} = \frac{1}{6} \ens_{1} + \frac{5}{6} \ens_{23456}$, where $\ens_{23456}$ represents a uniform distributions over the five outcomes, $1/6$ of $\ens$ can be constructed from $A$, but no more. Similarly, if $A_{12} = \{\ens_{1},\ens_{2}\}$, we can write $\ens = \frac{1}{3} \left(\frac{1}{2} \ens_1 + \frac{1}{2} \ens_2 \right)  + \frac{2}{3} \ens_{3456}$, so $1/3$ of $\ens$ can be constructed from $A$, but no more. Note that, given the uniform distribution, $1/6$ is exactly the probably for the event $A_1$ and $1/3$ the probability for event $A_{12}$. This gives the basic insight for our definition.

@@

One can then show that the fraction capacity $\frcap_{\ens} : \Sigma_{\Ens} \to [0,1]$ is a set function that satisfies the following:
\begin{enumerate}
	\item non-negative and unit bounded - $\frcap_{\ens}(A) \in [0,1]$
	\item monotone - $A \subseteq B \implies \frcap_{\ens}(A) \leq \frcap_{\ens}(B)$
	\item sub-additive - $\frcap_{\ens}(A \cup B) \leq \frcap_{\ens}(A) + \frcap_{\ens}(B)$
	\item continuous from below and above - $\frcap_{\ens}(\lim\limits_{i \to \infty} A_i) = \lim\limits_{i \to \infty} \frcap_{\ens}(A_i)$ for any increasing or decreasing sequence $\{A_i\}$.
\end{enumerate}

@@

If not for the lack of additivity over disjoint sets, the fraction capacity would be a probability measure. However, the difference between fraction capacity and probability goes beyond that. While the probability measure is defined over sets of outcomes, the fraction capacity is defined over ensembles, over probability distributions. Conceptually the two are different object, but they, as we will see later, linked. Suppose $\Ens$ is the classical ensemble space given by the set of probability distributions over a sample space $\Omega$. Let $U \subseteq \Omega$ be an event, let $A_U \subset \Ens$ be the set of all probability distributions with support within $U$ and $\ens \in \Ens$ a probability distribution. Then we can show that $\frcap_{\ens}() $
